---
title: "Covariates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, cache=T}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


We recommend installing the package using:
```{r install, eval=F}
devtools::install_github('rileymummah/SpFut.covariates', 
                         build_vignettes = T, force = T)
```


# Setup
First, let's load the library and the test data for the functions.

```{r setup}
library(SpFut.covariates)

data(locs)

head(locs)
```

`locs` contains two columns: grid.id and POLYGON geometry. For simplicity, let's only pull covariates for 2 grid locations. We also need to set a path to where the downloaded datasets are loaded and identify the column name of the grid cell ID number.

```{r}
locs <- locs[1:2,]
path <- "data/USA/"
id.label <- "grid.id"
```


## Population density
dat <- get_population(locs, id.label)

## Human footprint
dat <- get_footprint(locs, id.label)

## Traveltime
dat <- get_traveltime(locs, id.label)

## Soil composition
dat <- get_soil(locs, id.label)

## Climate
dat <- get_climate(locs, id.label)


# Download data

If you intend to use any of the following function, you must first download external datasets as instructed in the documentation.

## Landcover
dat <- get_landcover(locs, path, id.label)

## Elevation
dat <- get_elevation(locs, path, id.label)

## Protected areas
dat <- get_protected(locs, path, id.label)

## Flowlines
dat <- get_flowlines(locs, path, id.label)

## Water bodies
dat <- get_waterbodies(locs, path, id.label)

## Roads
dat <- get_roads(locs, path, id.label)
